Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-06-19 doc/network.tex
Editing and resolving comments

An epidemic can be characterized by the exponential growth rate, $r$, and the reproductive number, \RR.
The reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
\jd{Took out a comment about tracing and tried to clarify.}
However, estimating the reproductive number directly from disease natural history requires detailed knowledge which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, svensson2007note, roberts2007model, wallinga2007generation, park2019practical}.

@@

\jd{Still too much forward stuff for me; we don't use it for anything.}
\swp{Got rid of forward stuff and tried to clarify}
% Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
% As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles, 
% and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution.
Unlike the intrinsic generation-interval distribution, the shape of the observed generation-interval distributions depends on when and how they are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}. 
Here, we briefly describe how backward generation-interval distributions change over time because backward measurements better reflect real outbreak scenarios where contact tracing is performed by identifying infectors of infected individuals. (see \cite{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic} for discussions on forward generation intervals).
When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual, 
and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
When an epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and backward generation intervals will be longer than the intrinsic one \citep{nishiura2010time, tomba2010some, champredon2015intrinsic}.

In practice, generation intervals are often measured by aggregating available information from contact tracing. 
Whereas backward generation intervals represent infections that are one generation apart from a cohort of infectees, these aggregated generation intervals represent infections across multiple cohorts (i.e., the time of infection of the infectees can range anywhere from the beginning of an epidemic to the most recent infection).
\dc{I found the previous paragraph difficult to read and edited it to try to improve. The `cohorts` were not referring to the same ones (I think) and that was confusing. The bckwrd GI is observed at one given time and those infectees each had infectors from potentially different cohorts.}
\swp{I meant to say that because the backward GI is observed GI is observed at one given time, all infectees share the same infection time. Is this clearer?}
While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
These censored observations can be thought of as a weighted average of backward generation-interval distributions (weighted by incidence); \dc{weighted by what?} \swp{equation 8}
like them, the distribution of censored intervals will have a shorter mean than the intrinsic distribution early in an epidemic.

Realized generation intervals are also affected by spatial structure.
In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population even in the early phase of an epidemic. 
This means that contacts made late in an individual's infection are more likely to be ineffective due to infectious contacts that were made earlier (and so the contactee will have been infected already) 
\dc{Although I get you try to explain in simple terms what's going on with ``masked'', I think it's not really clear: masked by what?  is it that subsequent contacts in the neighbourhood do not count given they were infected earlier?}.
\swp{Tried to clarify}
As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
These observations allow us to make the following prediction: removing the censoring bias from observed generation intervals will give us a distribution that contains information about the population structure; using this distribution should allow us to correctly infer $\RR$ from $r$.

@@

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
An individual-level intrinsic infection kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
\dc{is it clearer to specify the kernel is about infections?}
\swp{Yes,}
Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness, which can depend both on the intrinsic infectiousness of a disease as well as the contact structure around an infected individual.

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations:
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
\swp{David seems to prefer da over dA. I think I prefer to keep it as dA to represent the space of variation? Need to clarify this?}
The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
\dc{wouldn't it be better to say: ``The population-level kernel describes the rate at which secondary infections are expected to be caused by an infected individual, on \emph{average}.'' I have troubles describing an ``average  infected individual''... Oh no, but you say it again below... well, I keep this comment, do whatever you want! }
\swp{I think that's clearer too.}

\dc{again, for me it's the intrinsic infectiousness averaged over the population, that's the $\int da$ thing. I stop from now on complaining about this ;)}
\swp{I think figure 1 shows that we're averaging across infectioussness as well as the time distribution. We also say above that ``Individuals may vary in their prognosis (e.g., duration of a latent and an infectious period) and their infectiousness''}
There are two components to an infection kernel: intrinsic infectiousness of an average infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
Assuming that a population mixes homogeneously, we can write: 
\begin{equation}
K(\tau) = \RR_0 g(\tau),
\end{equation}
where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population), 
and $g(\tau)$ is the intrinsic generation-interval distribution.

@@

When generation intervals are estimated through contact tracing during an outbreak, infection events that have not happened yet are not observed. 
This effect is called ``right-censoring''.
We can understand the effect of right-censoring using backward generation-interval distributions \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}:
the observed (censored) generation-interval distribution is a weighted average of backward generation-interval distributions (weighted by incidence) up until the observation time.
\dc{again, weighted by what? the binary variable ``observed/not observed''? It's disturbing that we talk about weights but never say what they are...}
\swp{They were (and are) in equation 8. I made it explicit.}

The density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
\begin{equation}
i_{t-\tau}(t) = \RR \, i(t-\tau) g(\tau) S(t)
\end{equation}
As shown in \cite{champredon2015intrinsic} \dc{I added the reference for the reader who wants to see how exactly this is derived}, the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
\begin{equation}\label{eq:backward}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-\sigma) g(\sigma) d\sigma}
\end{equation}
On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infected at time $s$ before $t$: $\int_0^t i_{s-\tau}(s) ds$.
Then, the censored generation-interval distribution is given by normalizing
\begin{equation}\label{eq:obsg}
\begin{aligned}
c_t(\tau) 
&= \frac{\int_0^t i_{s-\tau}(s) ds}{\int_0^t \int_0^t i_{s-\sigma}(s) ds d\sigma}= \frac{\int_0^t i(s-\tau) g(\tau) S(s) ds}{\int_0^t \int_0^t i(s-\sigma) g(x) S(s) ds d\sigma}\\
\end{aligned}
\end{equation}
Substituting \eref{backward} allows us to rewrite the censored generation-interval distribution as a weighted distribution of the backward generation intervals:
\begin{equation}
c_t(\tau) \propto \int_0^t b_s(\tau) i(s) ds.
\end{equation}

@@

\dc{shouldn't it be ``censored'' instead of ``observed mean''??}
\swp{I think observed in OK. Observation through contact tracing is what causes censoring.}
For a single outbreak, the observed mean  generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, any infection events that occur after the contact tracing period cannot be observed due to right censoring and short generation intervals are more likely to be observed.
In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the inverse exponentially weighted intrinsic generation-interval distribution \citep{britton2019estimation}:
\dc{notation $c_{exp}$ is not really consistent with Equation (9) $c_t$}
\jd{Use math formatting in comments or it can mess up the whole doc.}
\swp{Should I just write it as $c_t$? I can't come up with a better notation.}
\dc{what about $c_t^{\exp}$ ?}
\swp{TODO: ask JD}
\begin{equation}
c_t(\tau) = b_t(\tau) \propto g(\tau) \exp(-r\tau).
\label{eq:exp}
\end{equation}
A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
\dc{I'm confused... Figure~\ref{fig:censor} does not show they are the same. On the other hand, \autoref{eq:exp} is an equality... which one shall I believe?}
\swp{This is a notation issue... equality only holds during the exponential period. TODO: talk to Jonathan about notation.}
Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur, and the realized \dc{censored?} \swp{no, realized; there is no observation process in the second phenomenon} intervals will always be shorter \citep{champredon2015intrinsic}.
As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
Long backward generation intervals near the end of an epidemic have little effect on the censored generation-interval distributions because they represent only a small portion of an outbreak.
Overall, we expect, naively using the observed generation-interval distribution, to underestimate the reproductive number.

@@

The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
\end{equation}
\cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
\dc{but that is only during the exponential growth phase, isn't it?}
\swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.}
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the same result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{Champredon2018equivalence}:
\begin{equation}
g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma t} - e^{-\sigma t}\right),
\end{equation}
where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
\begin{equation}
\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)\tau} - e^{-\sigma \tau}\right)
\end{equation}
In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
In practice, directly using this distribution to link $r$ and $\RR$ is unrealistic because it requires that we know the per-pair contact rate. 
Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary meterial section 1.4.2);
we briefly describe this relationship in Methods.

@@

\jd{I want to keep thinking about network choice. Would a random network be useful? Are we careful enough about time windows (exponential phase vs (roughly) full single epidemic?}
\swp{Clarified time windows.}
\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible (simulated on a star network).
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or has recovered.
}
\label{fig:local}
\end{figure}

@@

Here, we investigate two methods for correcting for temporal bias in a contact tracing data (see Methods for details).
We refer to the first method (Section~\ref{poplevel}) as the population-level method as it relies on the observed distribution aggregated across the entire population.
When an epidemic is growing exponentially, the right-censoring causes the observed generation-interval to be inverse-exponentially weighted \eref{exp}; \dc{the previous sentence doesn't read well; forgot words?} \swp{Yes, forgot words. Not sure if ``inverse-exponentially weighted'' is clear, however.}
hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau),
\label{eq:geff}
\end{equation}
where $r$ is the exponential growth rate.
We refer to the second method (Section~\ref{indlevel}) as the individual-level method because it relies on individual contact information.
We model each infection as a non-homogeneous Poisson process arising from the infector \eref{nonhomogeneous}; 
incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.
For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$.

>>>>

An epidemic can be characterized by the exponential growth rate, $r$, and the reproductive number, \RR.
The reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
However, estimating the reproductive number directly from disease natural history requires detailed knowledge which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, svensson2007note, roberts2007model, wallinga2007generation, park2019practical}.

@@

% Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
% As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles, 
% and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution.
While the intrinsic generation-interval distribution is often assumed to be fixed, the shape of the observed generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}. 
Here, we briefly describe how backward generation-interval distributions change over time because backward measurements better reflect real outbreak scenarios where contact tracing is performed by identifying infectors of infected individuals. (see \cite{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic} for discussions on forward generation intervals).
When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual, 
and the distribution of backward generation intervals will be shorter than the intrinsic distribution \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
When an epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and backward generation intervals will be longer than the intrinsic one \citep{nishiura2010time, tomba2010some, champredon2015intrinsic}.

In practice, generation intervals are often measured by aggregating available information from contact tracing. 
%% Whereas backward generation intervals represent infections that are one generation apart from a cohort of infectees, these aggregated generation intervals represent infections across multiple cohorts (i.e., the time of infection of the infectees can range anywhere from the beginning of an epidemic to the most recent infection).
While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
Thus, there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
In fact, censored observations can be thought of as an average of backward generation-interval distributions, over the cohorts which have been infected before the observation time (and weighted by cohort size).
\jd{Simplified radically.}

Realized generation intervals are also affected by spatial structure.
In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population even in the early phase of an epidemic. 
This means that contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already) 
As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will give us a corrected distribution that contains information about the population structure; using this corrected distribution should allow us to correctly infer $\RR$ from $r$.

@@

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
An individual-level intrinsic infection kernel $K_a(\tau)$ describes the rate at which an infected individual $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
Individuals may vary in their course of infection (e.g., duration of latent and infectious periods) and their level of infectiousness, which can depend both on biological infectiousness as well contact patterns.

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations across an ``aspect space'' $A$ which describes the population distributions:
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
\swp{David seems to prefer da over dA. I think I prefer to keep it as dA to represent the space of variation? Need to clarify this?}
\jd{dA is better, and we obviously need to clarify.}
The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.

Assuming that a population mixes homogeneously, we can write: 
\begin{equation}
K(\tau) = \RR_0 g(\tau),
\end{equation}
where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population), 
and $g(\tau)$ is the expected time distribution of those cases (the intrinsic generation-interval distribution).

@@

When generation intervals are estimated through contact tracing during an outbreak, infection events that have not happened yet are not observed. 
This effect is called ``right-censoring''.
We can understand the effect of right-censoring using backward generation-interval distributions \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}:
the observed (censored) generation-interval distribution is a weighted average of backward generation-interval distributions (weighted by incidence) up until the observation time.

The density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
\begin{equation}
i_{t-\tau}(t) = \RR \, i(t-\tau) g(\tau) S(t)
\end{equation}
As shown in \cite{champredon2015intrinsic},
%% \dc{I added the reference for the reader who wants to see how exactly this is derived}
the backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occurred $\tau$ time units before the reference time $t$ and is proportional to $i(t-\tau) g(\tau)$:
\begin{equation}\label{eq:backward}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-\sigma) g(\sigma) d\sigma}
\end{equation}
On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any cohort infected at time $s$ before $t$: $\int_0^t i_{s-\tau}(s) ds$.
Then, the censored generation-interval distribution is given by normalizing
\begin{equation}\label{eq:obsg}
\begin{aligned}
c_t(\tau) 
&= \frac{\int_0^t i_{s-\tau}(s) ds}{\int_0^t \int_0^t i_{s-\sigma}(s) ds d\sigma}= \frac{\int_0^t i(s-\tau) g(\tau) S(s) ds}{\int_0^t \int_0^t i(s-\sigma) g(x) S(s) ds d\sigma}\\
\end{aligned}
\end{equation}
Substituting \eref{backward} allows us to rewrite the censored generation-interval distribution as a weighted distribution of the backward generation intervals:
\begin{equation}
c_t(\tau) \propto \int_0^t b_s(\tau) i(s) ds.
\end{equation}

@@

For a single outbreak, the observed mean  generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, longer generation intervals are more likely to be missed due to right censoring (and short generation intervals are more likely to be observed).
In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the intrinsic generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
\dc{notation $c_{exp}$ is not really consistent with Equation (9) $c_t$}
\swp{Should I just write it as $c_t$? I can't come up with a better notation.}
\dc{what about $c_t^{\exp}$ ?}
\swp{TODO: ask JD}
\jd{I like $b_0$ for the initial backward. We can then use $c_0$ for the initial censored, I guess.}
\begin{equation}
c_t(\tau) = b_t(\tau) \propto g(\tau) \exp(-r\tau).
\label{eq:exp}
\end{equation}
A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
\dc{I'm confused... Figure~\ref{fig:censor} does not show they are the same. On the other hand, \autoref{eq:exp} is an equality... which one shall I believe?}
\swp{This is a notation issue... equality only holds during the exponential period. TODO: talk to Jonathan about notation.}
Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
Long backward generation intervals near the end of an epidemic have little effect on the censored generation-interval distributions because they represent only a small portion of an outbreak.
Overall, we expect, naively using the observed generation-interval distribution, to underestimate the reproductive number.

@@

The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
\end{equation}
\cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
\dc{but that is only during the exponential growth phase, isn't it?}
\swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.}
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{Champredon2018equivalence}:
\begin{equation}
g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma t} - e^{-\sigma t}\right),
\end{equation}
where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
\begin{equation}
\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)\tau} - e^{-\sigma \tau}\right)
\end{equation}
In this case, with fixed infectiousness during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate): infecting a susceptible contact is analogous to no longer being infectious (since the contact cannot be infected again); therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
In practice, directly using this distribution to link $r$ and $\RR$ is unrealistic because it requires that we know the per-pair contact rate. 
Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary material section 1.4.2);
we briefly describe this relationship in Methods.

@@

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible (simulated on a star network).
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or has recovered.
}
\label{fig:local}
\end{figure}

@@

Here, we investigate two methods for correcting for temporal bias in a contact tracing data (see Methods for details).
We refer to the first method (Section~\ref{poplevel}) as the population-level method as it relies on the observed distribution aggregated across the entire population.
When an epidemic is growing exponentially, right-censoring causes the observed generation interval to be discounted by the exponential growth rate \eref{exp};
\swp{Forgot words. Not sure if ``inverse-exponentially weighted'' is clear, however.}
\jd{I fiddled the words -- but the equation ref is the key here; I think it's fine.}
hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau),
\label{eq:geff}
\end{equation}
where $r$ is the exponential growth rate.

We refer to the second method (Section~\ref{indlevel}) as the individual-level method because it relies on individual contact information.
We model each infection as a non-homogeneous Poisson process arising from the infector \eref{nonhomogeneous}; 
incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the censoring process in the observed generation intervals.
For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\RR$ from the observed growth rate $r$.
